---
title: "Emerging trend indicators of raccoon and other species in Europe"
editor: visual
project:
  type: website
  output-dir: docs
  execute-dir: project
format:
  html:
    df-print: paged
    toc: true
    toc-float: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-tools: true
execute:
  eval: true
  echo: true
  warning: true
  error: false
  include: true
from: markdown+emoji
---

## Introduction

This document shows an example of how using GBIF species occurrence cubes to assess the emerging status of some (alien) species in Europe and at country level. This workflow is strongly based on [**occurrence TrIAS indicators**](https://github.com/trias-project/indicators).

### Setup

First, list and load the needed packages.

```{r pkgs, message=FALSE, warning=FALSE}
library(readr)
library(here)
library(dplyr)
library(patchwork)
library(tidyr)
library(purrr)
library(tibble)
library(trias)
```

## Scope

### Taxonomic scope

The main species of interest is the raccoon (_Procyon lotor (Linnaeus, 1758)_, GBIF Key: [5218786](https://www.gbif.org/species/5218786)). We will also consider other species:
- [*Xenopus laevis*](https://www.gbif.org/species/5217334) (GBIF key: `5217334`)

The list could be extended in the future.

```{r define-species-of-interest}
species <- tibble::tibble(
  specieskey = c(5218786,
                 5217334),
  canonical_name = c("Procyon lotor",
                     "Xenopus laevis")
)
```

### Spatial scope

We are interested over the emerging status of the four species in **Europe**. A facetting at country level will follow later.

### Temporal scope

We request a species occurrence cube based on data from **1950**.

## Species occurrence cube

We triggered a GBIF occurrence cube via the [Occurrence SQL Download API](https://techdocs.gbif.org/en/data-use/api-sql-downloads) and on the hand of a JSON query ([query_cube_raccoon.json](../../data/input/query_cube_raccoon.json)). The resulting cube (DOI: [10.15468/dl.rx7qvb](https://doi.org/10.15468/dl.rx7qvb), `downloadKey`: `0011427-250127130748423`) can be downloaded in TSV format from GBIF. We have it saved at `data/input` as [`0011427-250127130748423.csv`](../../data/input/0011427-250127130748423.csv):

```{r read-cube, message=FALSE}
cube <- readr::read_tsv(
  here::here(
    "data",
    "input",
    "0011427-250127130748423.csv"
  )
)
```

Preview:

```{r}
head(cube)
```

Notice the presence of column `countryCode` as we grouped by country.

## From cubes to emerging status

### Preprocess: from cube to time series

We assess the emerging status of the species in Europe for 2024. First, we create time series up to 2024.

```{r}
last_year <- 2023
```

For each species, define cells with at least one observation:

```{r}
df_cc <-   cube |>
  dplyr::group_by(specieskey) |>
  dplyr::distinct(eeacellcode) |>
  dplyr::ungroup()
```

For each species, identify the first year with at least one observation:

```{r}
df_begin_year <-
  cube |>
  dplyr::group_by(specieskey) |>
  dplyr::summarize(begin_year = min(year))
```

For each species, combine `begin_year` and unique `eeacellcode` as found above:

```{r}
df_cc <-   df_cc |>
  dplyr::left_join(df_begin_year, by = "specieskey") |> 
  dplyr::select(specieskey, begin_year, eeacellcode)
```

Preview:

```{r}
head(df_cc)
```

For each cell (`eeacellcode`) and species (`specieskey`) we can now create a time series:

```{r}
# Define help function
make_time_series <- function(eeacellcode, specieskey, begin_year, last_year) {
  tidyr::expand_grid(
    eeacellcode = eeacellcode,
    specieskey = specieskey,
    year = seq(from = begin_year, to = last_year)
  )
}

# Create timeseries slots
df_ts <- purrr::pmap_dfr(df_cc,
  .f = make_time_series,
  last_year = last_year
)

# Add occurrence data
df_ts <-
  df_ts |>
  dplyr::left_join(
    cube |> dplyr::select(
      specieskey,
      year,
      eeacellcode,
      occurrences
    ),
    by = c("specieskey", "year", "eeacellcode")
  )

# Replace NAs with 0
df_ts <-
  df_ts |>
  tidyr::replace_na(list(occurrences = 0))
```

Add column for presence (1) or absence (0):

```{r}
df_ts <-
  df_ts |>
  dplyr::mutate(
    ispresent = dplyr::if_else(occurrences > 0, 1, 0)
  )
```

Save the time series as an interim output:

```{r save-time-series, message: false}
readr::write_tsv(
  df_ts,
  here::here("data", "interim", "time_series.tsv"),
  na = ""
)
```
